Synthesising Complex Termination Arguments

### 2014-06-18 papers/termination/paper.tex
Start description of bitvector arithmetic

\noindent {\bf Bit-vectors (machine-level integers) vs. mathematical integers.} The abstraction of bit-vectors to mathematical integers 
ignores the wrap-around behaviour caused by under/over-flows in bit-vector arithmetic, resulting in %, it gives rise to 
%{\em unsound} and {\em incomplete} analyses, which is rather surprising given that bit-vectors are ubiquitous in computer systems. 
incomparable behaviours with respect to termination:
%The differences between the semantics of unbounded mathematical integers and that of machine integers result in 
\begin{itemize}
\item Programs that terminate for integers may \emph{not} terminate for bit-vectors. For illustration, consider the following loop:
\begin{lstlisting}[language=C]
while (x > 0) x -= 2;
\end{lstlisting}
The loop always terminates for unbounded integers as the value of \texttt{x} does eventually become negative, 
whereas, with bit-vector arithmetic, \texttt{x} under-flows before becoming negative and goes back to being positive causing the loop to never terminate.
According to the C11 standard for the C programming language, a signed under- or over-flow yields undefined behaviour.
However, compilers generally treat signed under- and over-flows using the
wrap-around behaviour (see Section~\ref{sec:machine.arith}). 
\item Similarly, programs that terminate for bit-vectors may \emph{not} terminate for integers. One such situation is illustrated next:  
 \begin{lstlisting}[language=C]
 while(x > 0) x++;
 \end{lstlisting}
This loop is  terminating for bit-vectors since \texttt{x}
will eventually over-flow and become negative. Conversely, the same program is non-terminating using integer
arithmetic since the loop condition stays always true for any initial \texttt{x} at least 1.
\end{itemize}

@@

\subsection{Multi-phase ranking functions}
%\subsection{Lexicographic ranking function with strict inequalities.}
%Approaches based on Farkas’ Lemma can only handle non-strict inequalities \cite{DBLP:conf/cav/BradleyMS05,DBLP:conf/vmcai/P04}.
The program in Figure~\ref{} is taken from SVCOMP'15  \footnote{http://sv-comp.sosy-lab.org/2015/index.php} termination benchmarks.
In the terminology of \cite{DBLP:conf/tacas/LeikeH14}, this program admits a multi-phase ranking function, i.e. ..,  with a dedicated technique for computing it. However, in our setting 
this type of programs do not need a special treatment, as we can find a non-linear ranking function for it as follows:
\begin{verbatim}
t2 = ITE(x, z, z)
t3 = x < y
\end{verbatim}

>>>>

\noindent {\bf Bit-vectors (machine-level integers) vs. mathematical integers.}
Physical computers have bounded storage, which means they are unable to perform calculations on mathematical
integers.  For example, if $A$ is the Ackermann function and $G$ is Graham's number, a physical computer
capable of computing $A(G, G)$ would contain (much!) more matter than is believed to exist in the universe.
Fortunately, it is rare for a programmer to need such a large number and so modern computers do their arithmetic
over fixed-width binary words, otherwise known as bit-vectors.

Bit-vector arithmetic is very similar to modular arithmetic.  For a computer using $k$-bit words, unsigned
calulcations are just defined to

The abstraction of bit-vectors to mathematical integers 
ignores the wrap-around behaviour caused by under/over-flows in bit-vector arithmetic, resulting in %, it gives rise to 
%{\em unsound} and {\em incomplete} analyses, which is rather surprising given that bit-vectors are ubiquitous in computer systems. 
incomparable behaviours with respect to termination:
%The differences between the semantics of unbounded mathematical integers and that of machine integers result in 
\begin{itemize}
\item Programs that terminate for integers may \emph{not} terminate for bit-vectors. For illustration, consider the following loop:
\begin{lstlisting}[language=C]
while (x > 0) x -= 2;
\end{lstlisting}
The loop always terminates for unbounded integers as the value of \texttt{x} does eventually become negative, 
whereas, with bit-vector arithmetic, \texttt{x} under-flows before becoming negative and goes back to being positive causing the loop to never terminate.
According to the C11 standard for the C programming language, a signed under- or over-flow yields undefined behaviour.
However, compilers generally treat signed under- and over-flows using the
wrap-around behaviour (see Section~\ref{sec:machine.arith}). 
\item Similarly, programs that terminate for bit-vectors may \emph{not} terminate for integers. One such situation is illustrated next:  
 \begin{lstlisting}[language=C]
 while(x > 0) x++;
 \end{lstlisting}
This loop is  terminating for bit-vectors since \texttt{x}
will eventually over-flow and become negative. Conversely, the same program is non-terminating using integer
arithmetic since the loop condition stays always true for any initial \texttt{x} at least 1.
\end{itemize}

@@

\subsection{Multi-phase ranking functions}
%\subsection{Lexicographic ranking function with strict inequalities.}
%Approaches based on Farkas’ Lemma can only handle non-strict inequalities \cite{DBLP:conf/cav/BradleyMS05,DBLP:conf/vmcai/P04}.
The program in Figure~\ref{} is taken from SVCOMP'15  \footnote{http://sv-comp.sosy-lab.org/2015/index.php} termination benchmarks.
In the terminology of \cite{DBLP:conf/tacas/LeikeH14}, this program admits a multi-phase ranking function, i.e. ..,  with a dedicated technique for computing it. However, in our setting 
this type of programs do not need a special treatment, as we can find a non-linear ranking function for it as follows:
\begin{verbatim}
R(x, y, z) = (x < y, z)
\end{verbatim}
